---
title: "Data Manipulation_pdf"
author: "Oumo David"
date: "`r Sys.Date()`"
output: pdf_document
---

This is a tutorial on data manipulation in R using the mtcars dataset. 
This example covers common data manipulation tasks such as filtering, sorting, grouping, summarizing, and creating new variables.

# Load the Dataset
The mtcars dataset is included in R by default. You can load it using the data() function.

View the first few rows of the dataset


```{r}
data("mtcars")

head(mtcars)

```

# Basic Data Manipulations

## 1. Filter Rows
Filter the dataset to include only cars with more than 6 cylinders.

```{r}
library(dplyr)

filtered_data <- mtcars %>%
  filter(cyl > 6)

head(filtered_data)

```

## 2. Select Columns
Select specific columns (e.g., mpg, cyl, hp).

```{r}
selected_data <- mtcars %>%
  select(mpg, cyl, hp)

head(selected_data)

```

## 3. Sort Data
Sort the dataset by mpg (miles per gallon) in descending order.

```{r}
sorted_data <- mtcars %>%
  arrange(desc(mpg))

head(sorted_data)

```

## 4. Create New Variables
Add a new column kmpl (kilometers per liter) by converting mpg to kilometers per liter (1 mpg ≈ 0.425 kmpl).

```{r}
mutated_data <- mtcars %>%
  mutate(kmpl = mpg * 0.425)

head(mutated_data)

```

## 5. Group and Summarize
Group the data by cyl (number of cylinders) and calculate the mean mpg for each group.

```{r}
summary_data <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg),
            max_hp = max(hp))

print(summary_data)

```

## 6. Rename Columns
Rename the mpg column to miles_per_gallon.

```{r}
renamed_data <- mtcars %>%
  rename(miles_per_gallon = mpg)

head(renamed_data)

```

## 7. Add Row Numbers
Add a new column with row numbers.

```{r}
row_numbered_data <- mtcars %>%
  mutate(row_num = row_number())

head(row_numbered_data)

```

# Advanced Data Manipulations

## 1. Merge with Another Dataset
Create a small dataset and merge it with mtcars.

Create a small dataset

```{r}
car_names <- data.frame(
  car = rownames(mtcars),
  brand = c("Toyota", "Ford", "Mazda", "Honda", "Datsun", "Chrysler", "AMC", "Pontiac", "Fiat", "Porsche", "Lotus", "Ford", "Chevrolet", "Cadillac", "Lincoln", "Dodge", "Merc", "Valiant", "Duster", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc", "Merc")
)

```

Merge with mtcars

```{r}
merged_data <- mtcars %>%
  mutate(car = rownames(mtcars)) %>%
  left_join(car_names, by = "car")

head(merged_data)

```

## 2. Pivot Data
Reshape the data from wide to long format using pivot_longer.

```{r}
library(tidyr)

long_data <- mtcars %>%
  pivot_longer(cols = c(mpg, hp, wt), 
               names_to = "metric", 
               values_to = "value")

head(long_data)

```

## 3. Handle Missing Data
Add some missing values and handle them.

Introduce missing values

```{r}
mtcars_with_na <- mtcars
mtcars_with_na[1:5, "mpg"] <- NA

```

Remove rows with missing values

```{r}
cleaned_data <- mtcars_with_na %>%
  drop_na(mpg)

head(cleaned_data)

```

# Data visualization

Data visualization is a crucial part of data analysis, and R provides powerful tools like ggplot2 and base R plotting functions to create insightful visualizations. 
We will demonstrate various types of visualizations using the mtcars dataset.

Load necessary libraries

```{r}
library(ggplot2)

```

## 1. Scatter Plot
Visualize the relationship between two numeric variables, such as mpg (miles per gallon) and wt (weight).
We shall make a Scatter plot using ggplot2

```{r, fig.cap="Figure 1: The relationship between the weight of the motor vehicle and the miles per gallon"}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Scatter Plot of MPG vs Weight",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  theme_minimal()

```

## 2. Box Plot
Compare the distribution of mpg across different numbers of cylinders (cyl).
Let us make a Box plot using ggplot2

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_boxplot() +
  labs(title = "Box Plot of MPG by Cylinders",
       x = "Number of Cylinders",
       y = "Miles per Gallon") +
  theme_minimal()

```


## 3. Bar Plot
Visualize the average mpg for each number of cylinders.
Here, we shall create a Bar plot using ggplot2

```{r}
avg_mpg <- mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))

ggplot(avg_mpg, aes(x = factor(cyl), y = mean_mpg, fill = factor(cyl))) +
  geom_bar(stat = "identity") +
  labs(title = "Average MPG by Cylinders",
       x = "Number of Cylinders",
       y = "Average Miles per Gallon") +
  theme_minimal()

```

## 4. Histogram
Visualize the distribution of mpg.
Let us create a Histogram using ggplot2

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "orange", color = "black") +
  labs(title = "Histogram of MPG",
       x = "Miles per Gallon",
       y = "Frequency") +
  theme_minimal()

```

## 5. Density Plot
Visualize the density distribution of mpg.
Let us make a Density plot using ggplot2

```{r}
ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  labs(title = "Density Plot of MPG",
       x = "Miles per Gallon",
       y = "Density") +
  theme_minimal()

```

## 6. Line Plot
Visualize the relationship between mpg and hp (horsepower) using a line plot.
Here, we shall make a Line plot using ggplot2

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_line(color = "red") +
  labs(title = "Line Plot of MPG vs Horsepower",
       x = "Horsepower",
       y = "Miles per Gallon") +
  theme_minimal()

```

## 7. Heatmap
Visualize correlations between numeric variables in the dataset.
Let us make a Heatmap using base R

```{r}
correlation_matrix <- cor(mtcars)
heatmap(correlation_matrix, 
        col = colorRampPalette(c("blue", "white", "red"))(100),
        main = "Correlation Heatmap of mtcars Dataset")

```

## 8. Faceted Plot
Compare mpg vs wt for different numbers of cylinders using facets.
Making a Faceted plot using ggplot2

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "purple") +
  facet_wrap(~ cyl) +
  labs(title = "Faceted Scatter Plot of MPG vs Weight by Cylinders",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  theme_minimal()

```

## 9. Violin Plot
Visualize the distribution of mpg across different numbers of cylinders.
Let us make a Violin plot using ggplot2

```{r}
ggplot(mtcars, aes(x = factor(cyl), y = mpg, fill = factor(cyl))) +
  geom_violin() +
  labs(title = "Violin Plot of MPG by Cylinders",
       x = "Number of Cylinders",
       y = "Miles per Gallon") +
  theme_minimal()

```

## 10. Pair Plot
Visualize pairwise relationships between numeric variables.
Making a Pair plot using base R

```{r}
pairs(mtcars[, 1:5], 
      main = "Pair Plot of mtcars Dataset",
      pch = 19, 
      col = "darkblue")

```

## 11. Custom Theme
Customize the theme of your plots for better aesthetics.

Custom theme

```{r}
custom_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )

```

Example plot with custom theme

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "green", size = 3) +
  labs(title = "Custom Themed Scatter Plot",
       x = "Weight (1000 lbs)",
       y = "Miles per Gallon") +
  custom_theme

```
