---
title: "DESCRIPTIVE STATISTICS"
author: "Oumo David"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

This is a tutorial on Descriptive statistics.
Descriptive statistics summarize and organize characteristics of a dataset, providing insights into its distribution, central tendency, and variability. 
R offers various built-in functions and packages to perform descriptive statistics efficiently. In this tutorial, we will explore different methods to compute descriptive statistics in R.

# 1. Getting Started with R

Before we begin, ensure that R is installed on your system. You can also use RStudio for an enhanced coding experience.

To start, load the required packages:
If not installed, install the following packages

install.packages("dplyr")

install.packages("psych")


Load the necessary packages

```{r}
library(dplyr)

library(psych)

```

# 2. Creating a Sample Dataset

You can create a sample dataset using R’s built-in data.frame() function:
Create a sample dataset
View the dataset

```{r}
data <- data.frame(
  ID = 1:10,
  Age = c(25, 30, 35, 40, 29, 33, 31, 27, 26, 32),
  Salary = c(40000, 50000, 60000, 70000, 45000, 55000, 52000, 47000, 43000, 51000)
)

print(data)

```

# 3. Measures of Central Tendency

Central tendency describes the center point of a dataset. Common measures include:

Mean (Average)

```{r}
mean_age <- mean(data$Age)

mean_salary <- mean(data$Salary)

```

Median (Middle Value)

```{r}
median_age <- median(data$Age)

median_salary <- median(data$Salary)

```

Mode (Most Frequent Value)

```{r}
mode_function <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

mode_age <- mode_function(data$Age)

```

Display results

- Mean age, Median age, Mode_age

- Mean salary, Median salary

```{r}
mean_age; median_age; mode_age

mean_salary; median_salary

```
# 4. Measures of Dispersion

Dispersion indicates how data values spread around the central value.

Standard Deviation

```{r}
sd_age <- sd(data$Age)

sd_salary <- sd(data$Salary)

```

Variance

```{r}
var_age <- var(data$Age)

var_salary <- var(data$Salary)

```

Range

```{r}
range_age <- range(data$Age)

range_salary <- range(data$Salary)

```

Interquartile Range (IQR)

```{r}
iqr_age <- IQR(data$Age)

iqr_salary <- IQR(data$Salary)

```


Display results

- sd_age, var_age, range_age, iqr_age

- sd_salary, var_salary, range_salary, iqr_salary

```{r}
sd_age; var_age; range_age; iqr_age

sd_salary; var_salary; range_salary; iqr_salary

```
# 5. Summary Statistics Using Built-in Functions

R provides built-in functions for quick summaries: Summary function

```{r}
summary(data)

```

# 6. Using psych Package for Enhanced Descriptive Statistics

The describe() function from the psych package provides detailed statistics:

```{r}
describe(data)

```

# 7. Visualizing Data

Visualizations help understand data distribution.

Histogram

```{r}
hist(data$Age, main="Age Distribution", xlab="Age", col="blue", breaks=5)

```

Boxplot

```{r}
boxplot(data$Salary, main="Salary Distribution", ylab="Salary", col="red")

```

Descriptive statistics provide essential insights into datasets before conducting further analysis. R offers various functions and packages to compute these statistics efficiently. By understanding central tendency, dispersion, and visualization techniques, you can effectively summarize and interpret data.

# 8.Descriptive statistics on the mtcars dataset

Now, let us look at the mtcars dataset to explore the descriptive statistics

Load built-in dataset
Assign it to a variable

```{r}
data("mtcars")

df <- mtcars

```

# 9. Overview of the Dataset
To understand the dataset structure.

View first few rows

```{R}
head(df)

```

Check the structure

```{R}
str(df)

```

Summary statistics

```{r}
summary(df)

```

# 10. Measures of Central Tendency
These include the mean, median, and mode.

Mean
Mean of miles per gallon

```{r}
mean(df$mpg)   

```

Median
Median of miles per gallon

```{r}
median(df$mpg) 

```

Mode (Custom function, since R has no built-in mode function)

```{r}
mode_func <- function(x) {
  uniq_vals <- unique(x)
  uniq_vals[which.max(tabulate(match(x, uniq_vals)))]
}

mode_func(df$mpg)

```

# 11. Measures of Dispersion
These include the range, variance, standard deviation, and interquartile range (IQR).

Range

```{r}
range(df$mpg)

```

Variance

```{r}
var(df$mpg)

```

Standard deviation

```{r}
sd(df$mpg)

```


Interquartile Range (IQR)

```{r}
IQR(df$mpg)

```

# 12. Skewness and Kurtosis
Skewness measures symmetry, and kurtosis measures the "tailedness" of the distribution.

Load necessary library

```{r}
library(moments)

```

Skewness

```{r}
skewness(df$mpg)

```

Kurtosis

```{r}
kurtosis(df$mpg)

```

# 13. Frequency Distribution (Categorical Data)
For categorical data, we use tables and proportions.

Convert a numerical variable into a categorical one

```{r}
df$cyl_factor <- as.factor(df$cyl)

```

Frequency table

```{r}
table(df$cyl_factor)

```


Proportion table

```{r}
prop.table(table(df$cyl_factor))

```

# 14. Visualizing Data
Histograms

```{r}
hist(df$mpg, main="Histogram of MPG", xlab="Miles per Gallon", col="blue")

```

Boxplot

```{r}
boxplot(df$mpg, main="Boxplot of MPG", ylab="Miles per Gallon", col="red")

```

Bar Chart (for categorical data)

```{r}
barplot(table(df$cyl_factor), main="Cylinder Count", col="green")

```

# 15. Summary Statistics for Multiple Columns
Summary statistics for all numerical variables

```{r}
summary(df)

```

Apply functions across multiple columns
Mean for all numeric columns

```{r}
sapply(df[, sapply(df, is.numeric)], mean)

```

# Conclusion

This tutorial introduced key descriptive statistics in R, including measures of central tendency, dispersion, skewness, and visualizations. These techniques help summarize and understand data effectively.