---
title: "HYPOTHESIS TESTING"
author: "Oumo David"
date: "`r Sys.Date()`"
output: 
  html_document
  toc: true
  toc_float: true
  code_folding: hide
---

# Hypothesis Testing Using R

Hypothesis testing is a fundamental statistical technique used to make inferences about population parameters based on sample data. In this tutorial, we will explore how to perform hypothesis testing in R using common statistical tests, such as the t-test, chi-square test, and ANOVA.

# Understanding Hypothesis Testing

Hypothesis testing involves the following steps:

## 1. Formulate Hypotheses:

- Null Hypothesis (H₀): The default assumption (e.g., no effect or no difference).

- Alternative Hypothesis (H₁): The claim we want to test (e.g., there is an effect or difference).

## 2. Choose a Significance Level (α): Commonly set to 0.05 (5%).

## 3. Calculate the Test Statistic: A value computed from the sample data.

## 4. Determine the p-value: The probability of observing the test statistic under the null hypothesis.

## Make a Decision:

- If p-value < α, reject the null hypothesis.

- If p-value ≥ α, fail to reject the null hypothesis.

# Load Required Libraries

For this tutorial, we will use base R functions, so no additional libraries are required. However, we will use the ggplot2 package for visualization:

Install and load ggplot2 (for visualization)

install.packages("ggplot2")

```{R}
library(ggplot2)

```

# One-Sample t-Test

A one-sample t-test is used to compare the mean of a sample to a known value.

Example:

Suppose we have a sample of 30 students' test scores, and we want to test if the mean score is significantly different from 70.

- Sample data

```{R}
set.seed(123)
scores <- rnorm(30, mean = 75, sd = 10)

```

- One-sample t-test

```{R}
t_test_result <- t.test(scores, mu = 70)
t_test_result

```

## Interpretation:

- Null Hypothesis (H₀): The mean score is 70.

- Alternative Hypothesis (H₁): The mean score is not 70.

- If the p-value < 0.05, we reject H₀ and conclude that the mean score is significantly different from 70.

Since the p-value (0.01715) is less than the significance level (α = 0.05), we reject the null hypothesis. This means there is sufficient evidence to conclude that the sample mean is significantly different from 70.

Additionally, the 95% confidence interval (70.86573 to 78.19219) does not include the hypothesized mean (70), further supporting the conclusion that the true population mean is not 70.

# Two-Sample t-Test

A two-sample t-test is used to compare the means of two independent groups.

Example:

Suppose we have test scores for two groups of students (Group A and Group B) and want to test if their means are significantly different.

- Sample data

```{r}
set.seed(123)
group_a <- rnorm(30, mean = 75, sd = 10)
group_b <- rnorm(30, mean = 80, sd = 10)

```

- Two-sample t-test

```{r}
t_test_result <- t.test(group_a, group_b)
t_test_result

```

## Interpretation:

- Null Hypothesis (H₀): The means of Group A and Group B are equal.

- Alternative Hypothesis (H₁): The means of Group A and Group B are not equal.

- If the p-value < 0.05, we reject H₀ and conclude that the means are significantly different.

Since the p-value (0.003156) is less than the significance level (α = 0.05), we reject the null hypothesis. This means there is sufficient evidence to conclude that the means of the two groups are significantly different.

Additionally, the 95% confidence interval (-11.965426 to -2.543416) does not include zero, further supporting the conclusion that the true difference in means is not zero.

# Paired t-Test

A paired t-test is used to compare the means of two related groups (e.g., before and after measurements).

Example:

Suppose we have blood pressure measurements for 20 patients before and after a treatment.

- Sample data

```{r}
set.seed(123)
before <- rnorm(20, mean = 120, sd = 10)
after <- rnorm(20, mean = 115, sd = 10)

```

- Paired t-test

```{r}
t_test_result <- t.test(before, after, paired = TRUE)
t_test_result

```

## Interpretation:

- Null Hypothesis (H₀): The mean difference between before and after measurements is zero.

- Alternative Hypothesis (H₁): The mean difference is not zero.

- If the p-value < 0.05, we reject H₀ and conclude that the treatment had a significant effect.

Since the p-value (0.03159) is less than the significance level (α = 0.05), we reject the null hypothesis. This means there is sufficient evidence to conclude that the mean difference between the paired observations is significantly different from zero.

Additionally, the 95% confidence interval (0.679524 to 13.178095) does not include zero, further supporting the conclusion that the true mean difference is not zero.

# Chi-Square Test
A chi-square test is used to test the association between categorical variables.

Example:

Suppose we have a contingency table showing the distribution of gender (Male, Female) and preference for a product (Like, Dislike).

- Contingency table

```{r}
data <- matrix(c(30, 10, 20, 40), nrow = 2, byrow = TRUE)
rownames(data) <- c("Male", "Female")
colnames(data) <- c("Like", "Dislike")

```

- Chi-square test

```{r}
chi_test_result <- chisq.test(data)
chi_test_result

```

## Interpretation:

- Null Hypothesis (H₀): There is no association between gender and product preference.

- Alternative Hypothesis (H₁): There is an association between gender and product preference.

- If the p-value < 0.05, we reject H₀ and conclude that there is a significant association.

Since the p-value (0.0001052) is less than the significance level (α = 0.05), we reject the null hypothesis. This means there is sufficient evidence to conclude that there is a significant association between the two categorical variables.

# ANOVA (Analysis of Variance)

ANOVA is used to compare the means of three or more groups.

Example:

Suppose we have test scores for three groups of students (Group A, Group B, Group C) and want to test if their means are significantly different.

- Sample data

```{r}
set.seed(123)
group_a <- rnorm(30, mean = 75, sd = 10)
group_b <- rnorm(30, mean = 80, sd = 10)
group_c <- rnorm(30, mean = 85, sd = 10)

```

- Combine data into a data frame

```{r}
data <- data.frame(
    score = c(group_a, group_b, group_c),
    group = factor(rep(c("A", "B", "C"), each = 30))
)

```

## ANOVA

```{r}
anova_result <- aov(score ~ group, data = data)
summary(anova_result)

```

## Interpretation:

- Null Hypothesis (H₀): The means of all groups are equal.

- Alternative Hypothesis (H₁): At least one group mean is different.

- If the p-value < 0.05, we reject H₀ and conclude that there are significant differences between the groups.

Since the p-value (0.0000494) is less than the significance level (α = 0.05), we reject the null hypothesis. This means there is sufficient evidence to conclude that at least one group mean is significantly different from the others.

# Post-Hoc Analysis (Tukey's HSD)

If ANOVA results are significant, perform post-hoc analysis to identify which groups differ.

# Tukey's HSD test

```{r}
tukey_result <- TukeyHSD(anova_result)
tukey_result

```

## Interpretation:

- The output shows pairwise comparisons between groups and their p-values.

- Identify pairs with p-values < 0.05 as significantly different.

Significant Differences:

- B vs. A: There is a statistically significant difference between group B and group A (p < 0.05).

- C vs. A: There is a highly statistically significant difference between group C and group A (p < 0.001).

No Significant Difference:

- C vs. B: There is no statistically significant difference between group C and group B (p > 0.05).

# Visualization
Visualize the results using boxplots or bar plots.

Boxplot for ANOVA:

```{r}
ggplot(data, aes(x = group, y = score, fill = group)) +
    geom_boxplot() +
    theme_minimal() +
    labs(title = "Test Scores by Group", x = "Group", y = "Score")

```

# Summary

In this tutorial, we covered:

One-sample t-test: Compare a sample mean to a known value.

Two-sample t-test: Compare means of two independent groups.

Paired t-test: Compare means of two related groups.

Chi-square test: Test association between categorical variables.

ANOVA: Compare means of three or more groups.

Post-hoc analysis: Identify specific group differences after ANOVA.
